---
title: "Populations and Sampling Designs"
subtitle: "SD01 — Statistical Thinking"
author: "EulerLettersAI"
output:
  html_document:
    toc: true
format:
  html:
    code-links:
      - text: "Download R Markdown"
        icon: file-earmark-arrow-down
        href: SD01-03-populations-and-sampling.Rmd?download=1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

**Course:** Statistics and Probability for Data Science (DS14270).  
**Source deck:** Samer El Kababji, *SD01 — Statistical Thinking*.
Equations and teaching simulations below extend the deck's concepts; they are
not additional observations from the deck. Each file runs independently using
base R, with knitr/R Markdown needed only for rendering.

## Define the population and its target quantity

A sample is a subset of a population. The drug question in the deck targets
people with severe heart disease; it is not restricted to the one anecdotal
patient. For a computational sampling example, use a fictional league of
players, where teams are groups and salary is measured in thousands of dollars.
No real league salaries are used.

For a finite population of size $N$, its mean is

$$
\mu=\frac{1}{N}\sum_{i=1}^N y_i.
$$

**Worked example — a tiny fictional population.** If the entire population
has four salaries $y=(60,70,80,90)$ (in thousands of dollars), then

$$
\mu=\frac{60+70+80+90}{4}=75.
$$

The sum adds every population value. The simulation below uses a larger
fictional population.

```{r finite-population}
set.seed(1801)
team_sizes <- rep(40L, 10)
team <- rep(seq_along(team_sizes), team_sizes)
salary <- 50 + 10 * team + rgamma(length(team), shape = 4, scale = 5)
population <- data.frame(id = seq_along(team), team, salary)
N <- nrow(population)
mu <- mean(population$salary)
c(population_size = N, population_mean = mu)
```

This known population lets us compare samples with a truth that is usually
unknown in practice.

## Simple random sampling

A simple random sample of $n$ distinct people gives each person inclusion
probability $\pi_i=n/N$. Its mean estimates $\mu$:

$$
\bar y=\frac{1}{n}\sum_{i\in s}y_i,\qquad E(\bar y)=\mu.
$$

**Worked example — one sample.** From the four-player illustration above,
suppose the selected salaries are 60 and 80. The sum includes only these
sampled players:

$$
\bar y=\frac{60+80}{2}=70.
$$

One sample mean can differ from $\mu=75$; $E(\bar y)=\mu$ describes the
average over all possible simple random samples.

```{r simple-random-sample}
n <- 80
srs <- population[sample.int(N, n, replace = FALSE), ]
c(sample_size = nrow(srs), sample_mean = mean(srs$salary), population_mean = mu)
```

Under this design, the estimated standard error includes a finite-population
correction:

$$
\widehat{\operatorname{SE}}(\bar y)
=\sqrt{\left(1-\frac{n}{N}\right)\frac{s^2}{n}},\qquad
s^2=\frac{\sum_{i\in s}(y_i-\bar y)^2}{n-1}.
$$

**Worked example — the same sample.** With $N=4$, $n=2$, and
$\bar y=70$, expand the squared deviations first:

$$
s^2=\frac{(60-70)^2+(80-70)^2}{2-1}=200,
\qquad
\widehat{\operatorname{SE}}(\bar y)
=\sqrt{\left(1-\frac{2}{4}\right)\frac{200}{2}}
\approx7.07.
$$

The standard error is in thousands of dollars.

```{r srs-standard-error}
sqrt((1 - n / N) * var(srs$salary) / n)
```

## Stratified sampling: include every stratum

Treat team as a stratum and sample within every team. If stratum $h$ has
population size $N_h$, its weight is $W_h=N_h/N$. The estimator is

$$
\widehat\mu_{\mathrm{strat}}=\sum_{h=1}^H W_h\bar y_h.
$$

**Worked example — one stratified estimate.** In a separate small example,
suppose two teams contain 40 and 60 players, with sampled mean salaries of
70 and 90. Expand the sum into one weighted term per team:

$$
\widehat\mu_{\mathrm{strat}}
=\frac{40}{100}\times70+\frac{60}{100}\times90
=28+54=82.
$$

The weights come from population team sizes, and salaries are in thousands
of dollars.

```{r stratified-sample}
strata <- split(population, population$team)
stratified <- do.call(rbind, lapply(strata, function(d) {
  d[sample.int(nrow(d), 8, replace = FALSE), ]
}))
stratum_means <- tapply(stratified$salary, stratified$team, mean)
weights <- table(population$team) / N
stratified_mean <- sum(as.numeric(weights) * as.numeric(stratum_means))
c(sample_size = nrow(stratified), weighted_mean = stratified_mean)
```

With equal team sizes and eight sampled per team, the ordinary sample mean
also equals the weighted mean. With unequal team sizes and equal allocation,
use the population weights rather than averaging strata equally.

## Cluster sampling: include everyone in selected clusters

Now treat team as a cluster. Select $m$ teams from $H$ teams and include all
players on those teams. Each person's inclusion probability is $m/H$.
For known population size $N$, a design-unbiased estimator is

$$
\widehat\mu_{\mathrm{cluster}}
=\frac{H}{mN}\sum_{h\in C}\sum_{i\in h}y_i.
$$

**Worked example — one selected cluster.** In a separate tiny league with
$H=3$ teams of two players each, $N=6$. Suppose $m=1$ team is selected,
and its two salaries are 60 and 80. The inner sum adds salaries within that
team; the outer sum has only one selected team:

$$
\widehat\mu_{\mathrm{cluster}}
=\frac{3}{1\times6}(60+80)=70.
$$

```{r cluster-sample}
H <- length(strata)
m <- 2
selected_teams <- sample(seq_len(H), m, replace = FALSE)
cluster_sample <- population[population$team %in% selected_teams, ]
cluster_mean <- H / (m * N) * sum(cluster_sample$salary)
c(selected_teams = selected_teams, sample_size = nrow(cluster_sample),
  estimated_mean = cluster_mean)
```

Equal cluster sizes make this estimator equal to the sample mean here. Players
on one team share a salary component, so 80 clustered observations need not
contain as much information as 80 simple-random-sample observations. The SRS
standard-error formula above does not apply to this clustered design.

## Multistage sampling: sample within selected clusters

Select $m$ teams, then select $k$ players from each selected team of size $N_h$.
The person's inclusion probability and an inverse-probability estimator are

$$
\pi_{hi}=\frac{m}{H}\frac{k}{N_h},\qquad
\widehat\mu_{\mathrm{multi}}
=\frac{1}{N}\sum_{h\in C}\sum_{i\in s_h}\frac{y_{hi}}{\pi_{hi}}.
$$

**Worked example — one selected team.** Use the tiny six-player league
above, but now sample $k=1$ player from the selected team of size $N_h=2$.
If that player's salary is 80, then

$$
\pi_{hi}=\frac{1}{3}\times\frac{1}{2}=\frac{1}{6},
\qquad
\widehat\mu_{\mathrm{multi}}
=\frac{1}{6}\left(\frac{80}{1/6}\right)=80.
$$

Here both sums have one term. Dividing the observed salary by its inclusion
probability gives its estimated contribution to the population total, which
is then divided by $N$. The code below uses the larger simulated league.

```{r multistage-sample}
k <- 10
multistage <- do.call(rbind, lapply(selected_teams, function(h) {
  d <- strata[[as.character(h)]]
  sampled <- d[sample.int(nrow(d), k, replace = FALSE), ]
  sampled$inclusion_probability <- (m / H) * (k / nrow(d))
  sampled
}))
multistage_mean <- sum(multistage$salary / multistage$inclusion_probability) / N
c(sample_size = nrow(multistage), estimated_mean = multistage_mean)
```

## Convenience sampling can introduce bias

Taking players only from the first two teams gives zero inclusion probability
to everyone else. Repeating that procedure cannot remove its systematic bias.
For comparison, repeated SRS estimates fluctuate around the population mean:

$$
\operatorname{Bias}(\widehat\mu)=E(\widehat\mu)-\mu.
$$

```{r bias-versus-variation}
convenience <- population[population$team %in% c(1, 2), ]
repeated_srs <- replicate(1000,
  mean(population$salary[sample.int(N, 80, replace = FALSE)]))
c(population_mean = mu, convenience_mean = mean(convenience$salary),
  average_srs_mean = mean(repeated_srs),
  simulated_srs_bias = mean(repeated_srs) - mu)
hist(repeated_srs, breaks = 25, xlab = "Mean from an SRS of 80 players",
     main = "Sampling variation with a known population")
abline(v = mu, col = "steelblue", lwd = 2)
abline(v = mean(convenience$salary), col = "tomato", lwd = 2)
legend("topright", c("Population mean", "Convenience mean"),
       col = c("steelblue", "tomato"), lwd = 2)
```

Random selection reduces selection bias under the sampling design, but
nonresponse or incomplete population coverage can still bias the final data.

**Try it:** select only one cluster. Explain why its mean can differ greatly
from the population mean even though the cluster was selected randomly.
