---
title: "Data Matrices, Variable Types, and Association"
subtitle: "SD01 — Statistical Thinking"
author: "EulerLettersAI"
output:
  html_document:
    toc: true
format:
  html:
    code-links:
      - text: "Download R Markdown"
        icon: file-earmark-arrow-down
        href: SD01-02-data-matrices-and-variables.Rmd?download=1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

**Course:** Statistics and Probability for Data Science (DS14270).  
**Source deck:** Samer El Kababji, *SD01 — Statistical Thinking*.
Equations and teaching simulations below extend the deck's concepts; they are
not additional observations from the deck. Each file runs independently using
base R, with knitr/R Markdown needed only for rendering.

## Rows are observations; columns are variables

For $n$ observations and $p$ numeric variables, a data matrix has entries

$$
X=(x_{ij})\in\mathbb{R}^{n\times p},\qquad
x_{ij}=\text{value of variable }j\text{ for observation }i.
$$

An R data frame can also contain categorical columns. The next cell transcribes
the four loan records visible in the deck (original row labels 1, 2, 3, 50).
These are an excerpt, not the complete `loan50` dataset or a random sample.

```{r loan-excerpt}
loans <- data.frame(
  loan_amount = c(7500, 25000, 14500, 3000),
  interest_rate = c(7.34, 9.43, 6.08, 7.96),
  term = c(36L, 60L, 36L, 36L),
  grade = ordered(c("A", "B", "A", "A"), levels = LETTERS[1:7]),
  state = factor(c("MD", "OH", "MO", "CA")),
  total_income = c(70000, 254000, 80000, 34000),
  homeownership = factor(c("rent", "mortgage", "mortgage", "rent")),
  row.names = c("1", "2", "3", "50")
)
loans
loans["2", ]                 # one observational unit
loans$term                   # one variable
X <- as.matrix(loans[c("loan_amount", "interest_rate", "total_income")])
dim(X)
X[2, 1]                      # x_21: second displayed record, first variable
```

Do not convert the entire mixed-type data frame to a numeric matrix: factor
codes would not represent meaningful amounts or distances.

## Numeric, nominal, and ordinal variables

A discrete variable takes values in a countable set, such as a whole number of
months. A continuous measurement is modeled on an interval of real numbers,
even when recorded with rounding. Nominal categories have no inherent order;
ordinal categories have an order without necessarily equal spacing.

$$
\text{Term}\in\{1,2,3,\ldots\},\qquad
\text{Grade}\in\{A\prec B\prec\cdots\prec G\}.
$$

Here $\prec$ denotes category order, not a numerical gap. The statistical type
of a variable is distinct from its R storage class.

```{r variable-types}
data.frame(
  variable = names(loans),
  R_class = vapply(loans, function(z) paste(class(z), collapse = "/"), ""),
  statistical_type = c("numerical (amount)", "numerical (rate)",
                       "numerical discrete", "categorical ordinal",
                       "categorical nominal", "numerical (amount)",
                       "categorical nominal")
)
is.ordered(loans$grade)
levels(loans$homeownership)
```

## Summarize a column according to its type

For numeric values, the sample mean is $\bar x=n^{-1}\sum_i x_i$.
For category $c$, its relative frequency is
$\widehat p_c=n^{-1}\sum_i\mathbf{1}(x_i=c)$.

```{r column-summaries}
mean(loans$loan_amount)
prop.table(table(loans$homeownership))
```

These summarize only the four displayed records; they cannot characterize all
borrowers or all 50 loans.

## Association does not establish a causal link

For paired numeric variables, sample correlation is

$$
r=\frac{\sum_i(x_i-\bar x)(y_i-\bar y)}
{\sqrt{\sum_i(x_i-\bar x)^2\sum_i(y_i-\bar y)^2}}.
$$

The deck's sunscreen and ice cream example can arise from a shared cause:
temperature. The following fictional sales include small day-to-day changes
$u$ and $v$ in addition to the temperature effect:

$$
X=20+3T+u,\qquad Y=10+4T+v.
$$

Neither sales equation uses the other product's sales.

```{r shared-cause}
set.seed(1101)
temperature <- seq(15, 35, length.out = 100)
sunscreen_variation <- rep(c(-3, 2, 0, 4, -2), length.out = 100)
ice_cream_variation <- rep(c(2, -4, 3, -1, 0), length.out = 100)
sunscreen <- 20 + 3 * temperature + sunscreen_variation
ice_cream <- 10 + 4 * temperature + ice_cream_variation
x_centered <- sunscreen - mean(sunscreen)
y_centered <- ice_cream - mean(ice_cream)
r_formula <- sum(x_centered * y_centered) /
  sqrt(sum(x_centered^2) * sum(y_centered^2))
c(formula = r_formula, R_correlation = cor(sunscreen, ice_cream))
plot(sunscreen, ice_cream, pch = 19,
     xlab = "Simulated sunscreen sales", ylab = "Simulated ice cream sales")
```

A strong correlation alone cannot tell us whether changing sunscreen sales
would change ice cream sales. Also, zero correlation does not generally imply
independence; correlation describes linear association.
