---
title: "Descriptive Statistics and Quantiles"
subtitle: "SD02 — Exploratory Data Analysis"
author: "EulerLettersAI"
output:
  html_document:
    toc: true
format:
  html:
    code-links:
      - text: "Download R Markdown"
        icon: file-earmark-arrow-down
        href: SD02-01-descriptive-statistics-and-quantiles.Rmd?download=1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

**Course:** Statistics and Probability for Data Science (DS14270).  
**Source deck:** Samer El Kababji, *SD02 — Exploratory Data Analysis*.

This lesson uses small fictional datasets to connect each descriptive formula
to its base R implementation. Each code cell follows the mathematics it
implements.

## Population parameters and sample statistics

A **population** contains every observation of interest. A **sample** contains
the observations actually collected from that population. A numerical summary
of a population is called a **parameter**; a numerical summary of a sample is
called a **statistic**.

The same data values can be treated as a complete population in one question
and as a sample in another. The question determines which formula is suitable.

## Arithmetic mean

For observations $x_1,\ldots,x_n$, the sample mean is

$$
\bar{x}=\frac{1}{n}\sum_{i=1}^{n}x_i.
$$

**Worked example — eight deliveries.** For the values used below,

$$
\bar{x}=\frac{12+15+11+14+13+18+16+12}{8}
=\frac{111}{8}=13.875\text{ minutes}.
$$

The summation adds all eight delivery times; dividing by eight gives their
average.

```{r sample-mean}
delivery_minutes <- c(12, 15, 11, 14, 13, 18, 16, 12)
n <- length(delivery_minutes)

mean_by_formula <- sum(delivery_minutes) / n
mean_by_R <- mean(delivery_minutes)

c(formula = mean_by_formula, R = mean_by_R)
stopifnot(isTRUE(all.equal(mean_by_formula, mean_by_R)))
```

The mean uses every value. A particularly large or small observation can pull
it in that direction.

## Population and sample variance

Variance measures average squared distance from the mean. If the values are
the complete population of size $N$, population variance is

$$
\sigma^2=\frac{1}{N}\sum_{i=1}^{N}(x_i-\mu)^2.
$$

If the values are a sample of size $n$, sample variance is

$$
s^2=\frac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar{x})^2.
$$

**Worked example — the same three values, two interpretations.** Take
$x=(2,4,6)$, with mean $(2+4+6)/3=4$. The squared deviations sum to

$$
(2-4)^2+(4-4)^2+(6-4)^2=4+0+4=8.
$$

If these three values are the entire population, $\sigma^2=8/3\approx2.667$.
If they are a sample, $s^2=8/(3-1)=4$. Their sample standard deviation is
$s=\sqrt{4}=2$, in the original units. The denominator changes with the
population or sample interpretation.

Base R's `var()` computes sample variance, so its denominator is $n-1$.

```{r variance}
population_mean <- mean(delivery_minutes)
population_variance <- mean((delivery_minutes - population_mean)^2)

sample_mean <- mean(delivery_minutes)
sample_variance_formula <-
  sum((delivery_minutes - sample_mean)^2) / (n - 1)
sample_variance_R <- var(delivery_minutes)

c(population_variance = population_variance,
  sample_variance_formula = sample_variance_formula,
  sample_variance_R = sample_variance_R)
stopifnot(isTRUE(all.equal(sample_variance_formula, sample_variance_R)))
```

Squaring prevents positive and negative deviations from cancelling. Variance
is measured in squared units; the standard deviation `sd(delivery_minutes)` is
the square root of sample variance and returns to the original units.

## Proportion

Suppose $x$ of the $n$ sampled observations have a characteristic. The sample
proportion is

$$
\widehat p=\frac{x}{n}.
$$

**Worked example — defective items.** Two defective items among ten give
$\widehat p=2/10=0.20=20\%$. In indicator notation, the same calculation is
$(0+0+1+0+0+1+0+0+0+0)/10=0.20$.

In R, logical values behave like 1 for `TRUE` and 0 for `FALSE` when averaged.

```{r proportion}
defective <- c(FALSE, FALSE, TRUE, FALSE, FALSE,
               TRUE, FALSE, FALSE, FALSE, FALSE)

x <- sum(defective)
n_defects <- length(defective)
proportion_formula <- x / n_defects
proportion_R <- mean(defective)

c(defective_items = x, sample_size = n_defects,
  formula = proportion_formula, R = proportion_R)
```

## Median

First sort the values:

$$
x_{(1)}\leq x_{(2)}\leq\cdots\leq x_{(n)}.
$$

For odd $n$, the median is the middle ordered value. For even $n$, it is the
mean of the two middle ordered values.

**Worked example — find the middle after sorting.** For
$(2,4,7,10,12,15,18,20,25)$, $n=9$ and the middle position is
$(9+1)/2=5$, so the median is $x_{(5)}=12$. For $(3,5,8,9,14,21)$,
$n=6$ and the two middle positions are 3 and 4, giving $(8+9)/2=8.5$.

```{r median}
odd_values <- c(2, 4, 7, 10, 12, 15, 18, 20, 25)
even_values <- c(3, 5, 8, 9, 14, 21)

odd_sorted <- sort(odd_values)
even_sorted <- sort(even_values)

odd_median_formula <- odd_sorted[(length(odd_sorted) + 1) / 2]
middle_left <- length(even_sorted) / 2
even_median_formula <-
  (even_sorted[middle_left] + even_sorted[middle_left + 1]) / 2

c(odd_formula = odd_median_formula, odd_R = median(odd_values),
  even_formula = even_median_formula, even_R = median(even_values))
```

The median describes position after sorting. It is less affected by an extreme
value than the mean.

## Quartiles by linear interpolation

The standard quartiles correspond to

$$
Q_1=Q(0.25),\qquad Q_2=Q(0.50),\qquad Q_3=Q(0.75).
$$

For a target proportion $p$ between 0 and 1, the linear method used here finds
the fractional position

$$
h=1+(n-1)p.
$$

Write $h=j+g$, where $j=\lfloor h\rfloor$ and $0\leq g<1$. Then

$$
Q(p)=x_{(j)}+g\left(x_{(j+1)}-x_{(j)}\right).
$$

When $g=0$, the quantile is exactly the observed value $x_{(j)}$. This method
is `type = 7` in R's `quantile()` function.

**Worked example — interpolate the first quartile.** Use the six ordered
values $(3,5,8,9,14,21)$ and $p=0.25$:

$$
h=1+(6-1)(0.25)=2.25,\qquad j=2,\qquad g=0.25,
$$

$$
Q_1=x_{(2)}+0.25(x_{(3)}-x_{(2)})
=5+0.25(8-5)=5.75.
$$

The fractional position is one quarter of the way from the second value to
the third. Similarly, for $Q_3$, $h=1+5(0.75)=4.75$, so
$Q_3=9+0.75(14-9)=12.75$.

```{r linear-quantile-function}
linear_quantile <- function(values, p) {
  ordered <- sort(values)
  n <- length(ordered)
  h <- 1 + (n - 1) * p
  j <- floor(h)
  g <- h - j

  if (g == 0) {
    return(ordered[j])
  }

  ordered[j] + g * (ordered[j + 1] - ordered[j])
}

probabilities <- c(Q1 = 0.25, Q2 = 0.50, Q3 = 0.75)
by_formula <- vapply(probabilities,
                     function(p) linear_quantile(odd_values, p),
                     numeric(1))
by_R <- quantile(odd_values, probabilities, type = 7, names = FALSE)

rbind(formula = by_formula, R = by_R)
```

For the nine-value example, the positions are integers, giving
$Q_1=7$, $Q_2=12$, and $Q_3=18$.

For the six-value example, the median position is $h=3.5$, so it lies halfway
between the third and fourth ordered observations:

$$
Q_2=8+0.5(9-8)=8.5.
$$

```{r interpolated-quartiles}
data.frame(
  quartile = names(probabilities),
  p = probabilities,
  formula = vapply(probabilities,
                   function(p) linear_quantile(even_values, p),
                   numeric(1)),
  R = as.numeric(quantile(even_values, probabilities, type = 7))
)
```

R supports several quantile conventions. Always state the method when exact
quartile values matter. The linear method in this lesson is R's default
`type = 7`.

**Try it:** add a very large value such as 100 to `even_values`. Compare how
the mean and median change.
