---
title: "Comparing Variables with Plots"
subtitle: "SD02 — Exploratory Data Analysis"
author: "EulerLettersAI"
output:
  html_document:
    toc: true
format:
  html:
    code-links:
      - text: "Download R Markdown"
        icon: file-earmark-arrow-down
        href: SD02-03-comparing-variables-with-plots.Rmd?download=1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

**Course:** Statistics and Probability for Data Science (DS14270).  
**Source deck:** Samer El Kababji, *SD02 — Exploratory Data Analysis*.

Different plots answer different questions. This lesson uses base R and the
built-in `iris` dataset, together with one small fictional categorical dataset.

## Two numerical variables: scatter plot

For observation $i$, a scatter plot places one point at the coordinate

$$
(x_i,y_i).
$$

**Worked example — one flower, one point.** The first iris record has
sepal length 5.1 cm and petal length 1.4 cm. Its point is
$(x_1,y_1)=(5.1,1.4)$: move to 5.1 horizontally and 1.4 vertically.

The horizontal position represents one numerical variable and the vertical
position represents another. Here each point is one iris flower.

```{r scatter-plot}
head(iris[c("Sepal.Length", "Petal.Length")])

plot(iris$Sepal.Length, iris$Petal.Length,
     pch = 19, col = rgb(70, 130, 180, 140, maxColorValue = 255),
     xlab = "Sepal length (cm)", ylab = "Petal length (cm)",
     main = "Petal length versus sepal length")
```

Read a scatter plot by looking for direction, form, strength, clusters, gaps,
and unusual observations. The visible association does not by itself establish
that changing sepal length would cause petal length to change.

## Add a descriptive smooth

A smooth curve can help reveal the overall shape when many points overlap.
`lowess()` estimates local typical values using nearby observations. It is a
descriptive guide rather than a causal conclusion.

```{r scatter-with-smooth}
plot(iris$Sepal.Length, iris$Petal.Length,
     pch = 19, col = rgb(70, 130, 180, 140, maxColorValue = 255),
     xlab = "Sepal length (cm)", ylab = "Petal length (cm)",
     main = "Scatter plot with a descriptive smooth")
smooth_curve <- lowess(iris$Sepal.Length, iris$Petal.Length)
lines(smooth_curve, col = "tomato", lwd = 3)
```

The smooth summarizes the observed pattern. It should not hide clusters or
replace inspection of the original points.

## One categorical variable: counts and proportions

For category $c$, the frequency and sample proportion are

$$
n_c=\sum_{i=1}^{n}\mathbf{1}(z_i=c),\qquad
\widehat p_c=\frac{n_c}{n}.
$$

**Worked example — category indicators and proportions.** In the tiny
sequence (rent, own, rent, mortgage, rent), the indicators for rent are
$(1,0,1,0,1)$, so $n_{\text{rent}}=1+0+1+0+1=3$ and
$\widehat p_{\text{rent}}=3/5=0.60$. In the larger dataset below, the same
rule gives $38/100=0.38$ for rent, $47/100=0.47$ for mortgage, and
$15/100=0.15$ for own; the proportions sum to 1.

The following homeownership data are fictional.

```{r categorical-data}
homeownership <- factor(
  rep(c("rent", "mortgage", "own"), times = c(38, 47, 15)),
  levels = c("rent", "mortgage", "own")
)

category_counts <- table(homeownership)
category_proportions <- prop.table(category_counts)

category_counts
category_proportions
stopifnot(sum(category_counts) == length(homeownership),
          isTRUE(all.equal(sum(category_proportions), 1)))
```

A bar plot can display either counts or proportions. Bar plots leave space
between categories because the horizontal positions are labels, not a
continuous number line.

```{r categorical-bar-plots}
old_mfrow <- par("mfrow")
par(mfrow = c(1, 2))
barplot(category_counts, col = "steelblue", ylab = "Frequency",
        main = "Counts")
barplot(category_proportions, col = "steelblue", ylab = "Proportion",
        main = "Proportions")
par(mfrow = old_mfrow)
```

The plots have the same shape because every count was divided by the same
total. The vertical scales answer different questions.

## Numerical response across categories: box plots

To compare a numerical variable across groups, compute a separate five-number
summary for each category:

$$
(\text{minimum},Q_1,\text{median},Q_3,\text{maximum}).
$$

**Worked example — summarize one group.** Suppose one category has ordered
values $(3,5,8,9,14,21)$. With `type = 7` quartiles, its five-number summary is

$$
(3,\ 5.75,\ 8.5,\ 12.75,\ 21).
$$

The minimum and maximum are the endpoints, the median is $(8+9)/2=8.5$,
and the quartiles interpolate at positions 2.25 and 4.75. Repeat the
calculation separately for each category. Box-plot whiskers reach the most
extreme observations within their outlier fences; they need not equal the
minimum and maximum when outliers are present.

```{r grouped-summaries}
sepal_by_species <- split(iris$Sepal.Length, iris$Species)

group_summaries <- do.call(rbind, lapply(sepal_by_species, function(values) {
  c(minimum = min(values),
    Q1 = unname(quantile(values, 0.25, type = 7)),
    median = median(values),
    Q3 = unname(quantile(values, 0.75, type = 7)),
    maximum = max(values))
}))
group_summaries
```

```{r grouped-boxplots}
boxplot(Sepal.Length ~ Species, data = iris,
        col = c("lightblue", "lightgreen", "mistyrose"),
        xlab = "Species", ylab = "Sepal length (cm)",
        main = "Sepal length by iris species")
```

Box plots make medians, quartiles, and suspected outliers easy to compare, but
they hide many details of the shape inside each group.

## Box plots and violin plots

A violin plot uses a smooth density estimate $\widehat f_g(y)$ for each group
$g$. At height $y$, the violin width is proportional to
$\widehat f_g(y)$. Wider sections indicate values that occur more densely.

**Worked example — convert density to width.** The helper below scales each
violin's largest density to a half-width of 0.38:

$$
\text{half-width}(y)=0.38\,
\frac{\widehat f_g(y)}{\max_t\widehat f_g(t)}.
$$

If the maximum estimated density in one group is 0.8 and the density at a
particular height is 0.4, its half-width there is
$0.38(0.4/0.8)=0.19$. Around a category centered at 2, the left and right
edges are $2-0.19=1.81$ and $2+0.19=2.19$. These density values are
illustrative; each group's actual estimates come from `density()`.
Because each violin is scaled by its own maximum, equal widths across groups
do not necessarily mean equal absolute densities.

The helper below draws the violin shapes using base R's `density()` function.
The dot in each violin is the group median.

```{r boxplot-versus-violin}
draw_violin <- function(values, center, color, width = 0.38) {
  estimate <- density(values)
  half_width <- width * estimate$y / max(estimate$y)
  polygon(
    x = c(center - half_width, rev(center + half_width)),
    y = c(estimate$x, rev(estimate$x)),
    col = color,
    border = "grey30"
  )
  points(center, median(values), pch = 19)
}

species_names <- levels(iris$Species)
colors <- c("lightblue", "lightgreen", "mistyrose")
value_range <- range(iris$Sepal.Length)

old_mfrow <- par("mfrow")
par(mfrow = c(1, 2))

boxplot(Sepal.Length ~ Species, data = iris, col = colors, xaxt = "n",
        xlab = "Species", ylab = "Sepal length (cm)", main = "Box plots")
axis(1, at = seq_along(species_names), labels = species_names,
     cex.axis = 0.65)

plot(NA, xlim = c(0.5, 3.5), ylim = value_range, xaxt = "n",
     xlab = "Species", ylab = "Sepal length (cm)", main = "Violin plots")
axis(1, at = seq_along(species_names), labels = species_names,
     cex.axis = 0.65)
for (i in seq_along(species_names)) {
  draw_violin(sepal_by_species[[species_names[i]]], i, colors[i])
}

par(mfrow = old_mfrow)
```

The violin shows more of the internal shape, while the box plot gives a more
direct view of quartiles. A violin outline is smoothed, so small bumps should
not be overinterpreted, especially in small groups.

**Try it:** replace `Sepal.Length` with `Petal.Length` in the grouped summaries
and plots. Which species become easier to distinguish?
