---
title: "Logistic Regression and p-values"
subtitle: "A medical example with 30-day recovery"
author: "EulerLettersAI"
format:
  html:
    code-links:
      - text: "Download R Markdown"
        icon: file-earmark-arrow-down
        href: 1-logistic-regression.Rmd?download=1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Learning goals

By the end of this lesson, you should be able to:

1. explain why logistic regression is used for a yes/no outcome;
2. interpret coefficients as odds ratios;
3. explain what a p-value does—and does not—tell us;
4. use an effect estimate, confidence interval, and clinical context together;
5. recognize when clustered medical data need a mixed-effects model.

## The medical question

Imagine a fictional study of 300 patients in 10 clinics. Each patient received
either a new treatment (`treatment = 1`) or standard care (`treatment = 0`). The
outcome `recovered` is 1 if the patient recovered within 30 days and 0 otherwise.
We also record age because age may affect recovery.

Our main question is:

> After accounting for age, is the new treatment associated with 30-day recovery?

The data below are simulated for teaching. They do not describe real patients
or establish that any real treatment works.

### Mathematics: generate one binary outcome per patient

Let $Y_i$ denote recovery for patient $i$, $T_i$ denote treatment, and
$A_i=\text{Age}_i-60$ denote centered age. A binary outcome follows a Bernoulli
distribution:

$$
Y_i\mid p_i \sim \operatorname{Bernoulli}(p_i),
\qquad
P(Y_i=y_i)=p_i^{y_i}(1-p_i)^{1-y_i}.
$$

To make the same simulated data useful in Lesson 2, patients also have a clinic
effect $u_{j[i]}$. The data-generating equation is

$$
\eta_i=-0.40+\log(1.8)T_i-0.035A_i+u_{j[i]},
\qquad
u_j\sim N(0,0.65^2),
$$

and the inverse-logit function converts the linear predictor to a probability:

$$
p_i=\operatorname{logit}^{-1}(\eta_i)
=\frac{e^{\eta_i}}{1+e^{\eta_i}}.
$$

The R cell implements these equations. `plogis(eta)` is R's numerically stable
version of $\operatorname{logit}^{-1}(\eta)$.

```{r simulate-data}
set.seed(2026)

n_clinics <- 10
patients_per_clinic <- 30
n <- n_clinics * patients_per_clinic

clinic <- factor(rep(seq_len(n_clinics), each = patients_per_clinic))
clinic_effect <- rnorm(n_clinics, mean = 0, sd = 0.65)
treatment <- rbinom(n, size = 1, prob = 0.50)
age <- round(pmin(pmax(rnorm(n, mean = 60, sd = 11), 30), 85))

# The simulation uses a treatment log-odds effect of log(1.8).
linear_predictor <- -0.40 + log(1.8) * treatment - 0.035 * (age - 60) +
  clinic_effect[as.integer(clinic)]
probability_recovery <- plogis(linear_predictor)
recovered <- rbinom(n, size = 1, prob = probability_recovery)

medical <- data.frame(
  clinic = clinic,
  treatment = treatment,
  age = age,
  recovered = recovered
)

head(medical)
```

## Start with descriptive results

Before fitting a model, inspect the observed recovery proportions.

### Mathematics: an observed proportion

For treatment group $t\in\{0,1\}$, the empirical recovery proportion is

$$
\widehat p_t=\frac{1}{n_t}\sum_{i:T_i=t}Y_i.
$$

The next cell groups patients by $T_i$ and computes this sample mean.

```{r descriptive-results}
recovery_table <- with(
  medical,
  aggregate(recovered, list(treatment = treatment), mean)
)
names(recovery_table)[2] <- "recovery_proportion"
recovery_table$treatment <- c("Standard care", "New treatment")
recovery_table
```

This is an *unadjusted* comparison. It does not account for age, and it treats
all patients as independent even though patients from one clinic may resemble
one another.

## Why not ordinary linear regression?

For a binary outcome, a straight-line model can predict impossible values below
0 or above 1. Logistic regression instead models the **log-odds**:

$$
\log\left(\frac{p_i}{1-p_i}\right)
= \beta_0 + \beta_1\text{Treatment}_i
+ \beta_2(\text{Age}_i-60),
$$

where $p_i$ is patient $i$'s probability of recovery. Centering age at 60 makes
the intercept refer to a 60-year-old patient rather than a hypothetical
zero-year-old patient.

Because observations are Bernoulli, the likelihood for
$\boldsymbol\beta=(\beta_0,\beta_1,\beta_2)$ is

$$
L(\boldsymbol\beta)
=\prod_{i=1}^{n}p_i^{y_i}(1-p_i)^{1-y_i},
$$

or, more conveniently, the log-likelihood is

$$
\ell(\boldsymbol\beta)
=\sum_{i=1}^{n}\left[y_i\log(p_i)+(1-y_i)\log(1-p_i)\right].
$$

Logistic regression chooses the maximum-likelihood estimate
$\widehat{\boldsymbol\beta}=\arg\max_{\boldsymbol\beta}\ell(\boldsymbol\beta)$.
Unlike ordinary least squares, this optimization has no simple closed-form
solution, so `glm()` solves it iteratively.

Fit the model with `glm()` and the binomial family:

```{r fit-logistic-model}
medical$age_centered <- medical$age - 60

ordinary_model <- glm(
  recovered ~ treatment + age_centered,
  data = medical,
  family = binomial(link = "logit")
)

summary(ordinary_model)
```

## From coefficients to odds ratios

The coefficients printed by R are changes in log-odds. Exponentiating a
coefficient gives an **odds ratio (OR)**, which is usually easier to discuss.

For treatment, holding age fixed,

$$
\frac{\operatorname{odds}(Y=1\mid T=1,A)}
{\operatorname{odds}(Y=1\mid T=0,A)}
=\frac{e^{\beta_0+\beta_1+\beta_2A}}
{e^{\beta_0+\beta_2A}}
=e^{\beta_1}.
$$

A large-sample 95% Wald interval is first formed on the log-odds scale and then
exponentiated:

$$
\widehat\beta_k\pm1.96\,SE(\widehat\beta_k)
\quad\Longrightarrow\quad
\left[e^{\widehat\beta_k-1.96SE},
e^{\widehat\beta_k+1.96SE}\right].
$$

The next cell performs exactly these transformations for every coefficient.

```{r odds-ratios}
coefficient_table <- summary(ordinary_model)$coefficients

results <- data.frame(
  term = rownames(coefficient_table),
  estimate_log_odds = coefficient_table[, "Estimate"],
  odds_ratio = exp(coefficient_table[, "Estimate"]),
  confidence_low = exp(coefficient_table[, "Estimate"] -
                         1.96 * coefficient_table[, "Std. Error"]),
  confidence_high = exp(coefficient_table[, "Estimate"] +
                          1.96 * coefficient_table[, "Std. Error"]),
  p_value = coefficient_table[, "Pr(>|z|)"],
  row.names = NULL
)

results
```

Interpret the treatment row as follows:

- An OR of 1 means equal odds of recovery in the two treatment groups.
- An OR above 1 means higher odds with the new treatment.
- An OR below 1 means lower odds with the new treatment.
- For example, OR = 1.50 means 50% higher **odds**, not 50 percentage points
  higher probability and not necessarily 50% higher risk.

For age, the OR is for each additional year because age is measured in years.
An age OR of 0.96 would mean about 4% lower odds per additional year:
$100(0.96-1)=-4\%$. Centering age changes the intercept, but it does not change
the age slope, its odds ratio, or its p-value.

## Odds are not probabilities

The same odds ratio can correspond to different changes in probability,
depending on the starting risk. Predicted probabilities make the result more
concrete. Here are predictions for a 60-year-old patient:

For treatment value $t$ and centered age $a$, first calculate

$$
\widehat\eta(t,a)=\widehat\beta_0+\widehat\beta_1t+
\widehat\beta_2a,
$$

then transform it back from log-odds:

$$
\widehat p(t,a)=\frac{1}{1+e^{-\widehat\eta(t,a)}}.
$$

The code uses `type = "response"` to request this inverse-logit transformation
rather than the raw linear predictor.

```{r predicted-probabilities}
new_patients <- data.frame(
  treatment = c(0, 1),
  age_centered = c(0, 0)
)

new_patients$group <- c("Standard care", "New treatment")
new_patients$predicted_recovery <- predict(
  ordinary_model,
  newdata = new_patients,
  type = "response"
)
new_patients[, c("group", "predicted_recovery")]
```

These model-based probabilities are often more understandable to patients and
clinicians than odds ratios. They are estimates, not guarantees for an
individual patient.

## What exactly is a p-value?

For the treatment coefficient, the model tests:

$$
H_0: \beta_1=0 \quad \text{or equivalently} \quad H_0: OR=1.
$$

R reports a two-sided Wald p-value. Informally, it asks:

> If the treatment truly had no association with recovery, and the model's
> assumptions were correct, how surprising would an estimate at least this far
> from zero be because of random sampling variation?

A small p-value means the observed estimate is relatively incompatible with
that null model. It is **not**:

- the probability that the null hypothesis is true;
- the probability that the result happened "by chance";
- the probability that the treatment works;
- the size or clinical importance of the effect;
- proof that treatment causes recovery.

For example, `p = 0.03` does **not** mean there is a 3% probability that no
treatment effect exists. It describes the data under a null hypothesis; it does
not assign a probability to that hypothesis.

## How the p-value is calculated

For each coefficient, R divides the estimate by its standard error:

$$z=\frac{\widehat\beta}{SE(\widehat\beta)}.$$

It then compares $|z|$ with a standard normal distribution. We can reproduce
the treatment p-value:

$$
p\text{-value}=P(|Z|\ge |z_{obs}|\mid H_0)
=2\left[1-\Phi(|z_{obs}|)\right]
=2\Phi(-|z_{obs}|),
$$

where $Z\sim N(0,1)$ under the large-sample null approximation and $\Phi$ is
the standard normal cumulative distribution function. The R functions
`pnorm()` and `abs()` translate this equation directly.

```{r calculate-p-value}
treatment_estimate <- coef(ordinary_model)["treatment"]
treatment_se <- coefficient_table["treatment", "Std. Error"]
z_value <- treatment_estimate / treatment_se
p_value <- 2 * pnorm(-abs(z_value))

c(z_value = z_value, p_value = p_value)
```

The standard error reflects uncertainty. Larger samples generally produce
smaller standard errors, so a tiny effect can have a small p-value in a very
large study. Conversely, a clinically important estimated effect can have a
large p-value in a small, imprecise study.

## Reading p-values with confidence intervals

At a conventional two-sided 5% significance level, these statements usually
agree for the same coefficient and method:

- `p < 0.05`;
- the 95% confidence interval for the log-odds coefficient excludes 0;
- the 95% confidence interval for the odds ratio excludes 1.

The confidence interval is more informative because it shows both the estimated
effect and its precision. A wide interval means many effect sizes remain
compatible with the data.

Avoid saying "there is no effect" merely because `p >= 0.05`. A better statement
is: "The study did not provide strong evidence against no association; the
confidence interval shows which beneficial or harmful effects remain plausible."

## A responsible decision process

Do not let `p < 0.05` make the decision by itself. Before seeing the data, define
the primary outcome, model, significance level, and a **minimum clinically
important effect**. Then ask:

1. **What is the direction and size?** Read the odds ratio and predicted
   probabilities.
2. **How uncertain is it?** Read the confidence interval, not only the p-value.
3. **Is it clinically meaningful?** Compare the estimate and interval with a
   pre-specified clinical threshold.
4. **Is the design credible?** Consider randomization, missing data,
   confounding, measurement quality, and whether assumptions are reasonable.
5. **Were many hypotheses tested?** Selective reporting and multiple testing
   can create small p-values even when no real effects exist.
6. **Do benefits outweigh harms, costs, and patient preferences?** Statistical
   evidence is only one part of a medical decision.

A compact conclusion template is:

> After adjustment for age, the estimated treatment odds ratio was **[OR]**
> (95% CI **[lower, upper]**, p = **[p]**). This is [compatible/not compatible]
> with no association at the pre-specified 5% level. The estimated probability
> difference for a 60-year-old was **[difference]** percentage points. Clinical
> importance should be judged against the pre-specified threshold and study
> limitations.

Use the next chunk to fill in those values.

For the 60-year-old example, the model-based absolute effect is the probability
difference

$$
\widehat\Delta_p=\widehat p(T=1,A=0)-\widehat p(T=0,A=0).
$$

This is different from the odds ratio $e^{\widehat\beta_1}$ and is usually
reported in percentage points.

```{r conclusion-values}
treatment_result <- results[results$term == "treatment", ]
probability_difference <- diff(new_patients$predicted_recovery)

c(
  odds_ratio = treatment_result$odds_ratio,
  confidence_low = treatment_result$confidence_low,
  confidence_high = treatment_result$confidence_high,
  p_value = treatment_result$p_value,
  probability_difference = probability_difference
)
```

## An important limitation: patients are clustered

The ordinary model assumes that, after accounting for treatment and age, all
observations are independent. But patients treated in the same clinic may share
staff, procedures, referral patterns, or local populations. Their outcomes may
therefore be correlated.

Ignoring that correlation can make standard errors too small and p-values too
optimistic. The second lesson uses the same data and adds a clinic-level random
intercept.

## Check your understanding

1. If the treatment OR is 1.7, does that mean recovery probability rose by 70
   percentage points? Why not?
2. If p = 0.08, can you conclude that treatment has no effect?
3. Which communicates more information: a p-value alone, or an odds ratio with
   a confidence interval? Why?
4. What part of this study makes the independence assumption questionable?
