---
title: "Mixed-Effects Logistic Regression"
subtitle: "Accounting for patients grouped within clinics"
author: "EulerLettersAI"
format:
  html:
    code-links:
      - text: "Download R Markdown"
        icon: file-earmark-arrow-down
        href: 2-mixed-effects-logistic-regression.Rmd?download=1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Learning goals

By the end of this lesson, you should be able to:

1. explain why patients within a clinic may not be independent;
2. distinguish fixed effects from random effects;
3. fit a random-intercept logistic model;
4. interpret the treatment effect and clinic variation;
5. explain why a mixed model may change standard errors and p-values.

This lesson uses exactly the same fictional study and simulation as the first
lesson. The outcome is 30-day recovery, and 300 patients are nested in 10
clinics.

## Recreate the shared medical data

For patient $i$ in clinic $j$, the simulation follows

$$
Y_{ij}\mid p_{ij},u_j\sim\operatorname{Bernoulli}(p_{ij}),
\qquad u_j\sim N(0,0.65^2),
$$

$$
\operatorname{logit}(p_{ij})
=-0.40+\log(1.8)T_{ij}-0.035(Age_{ij}-60)+u_j.
$$

Using the same random seed and the same equations as Lesson 1 produces exactly
the same patients. The only change in this lesson is the model used to analyze
them.

```{r simulate-same-data}
set.seed(2026)

n_clinics <- 10
patients_per_clinic <- 30
n <- n_clinics * patients_per_clinic

clinic <- factor(rep(seq_len(n_clinics), each = patients_per_clinic))
clinic_effect <- rnorm(n_clinics, mean = 0, sd = 0.65)
treatment <- rbinom(n, size = 1, prob = 0.50)
age <- round(pmin(pmax(rnorm(n, mean = 60, sd = 11), 30), 85))
linear_predictor <- -0.40 + log(1.8) * treatment - 0.035 * (age - 60) +
  clinic_effect[as.integer(clinic)]
probability_recovery <- plogis(linear_predictor)
recovered <- rbinom(n, size = 1, prob = probability_recovery)

medical <- data.frame(
  clinic = clinic,
  treatment = treatment,
  age = age,
  age_centered = age - 60,
  recovered = recovered
)

head(medical)
```

## Why ordinary logistic regression may be insufficient

Patients in one clinic share an environment. Even after accounting for
treatment and age, two patients from the same clinic may be more alike than two
patients from different clinics. Ordinary logistic regression assumes the
remaining observations are independent, so its uncertainty calculation can be
misleading when outcomes are clustered.

Compare recovery proportions across clinics:

For clinic $j$, with $n_j$ patients, the observed clinic recovery proportion is

$$
\widehat p_j=\frac{1}{n_j}\sum_{i=1}^{n_j}Y_{ij}.
$$

The next cell evaluates $n_j$ and $\widehat p_j$ for all ten clinics.

```{r clinic-proportions}
clinic_summary <- aggregate(
  recovered ~ clinic,
  data = medical,
  FUN = function(x) c(n = length(x), recovery = mean(x))
)

clinic_summary <- data.frame(
  clinic = clinic_summary$clinic,
  n = clinic_summary$recovered[, "n"],
  recovery_proportion = clinic_summary$recovered[, "recovery"]
)
clinic_summary
```

Some variation is sampling noise, and some may reflect real clinic-to-clinic
differences. A mixed-effects model estimates the amount of extra variation.

## The random-intercept model

We extend the earlier equation:

$$
\log\left(\frac{p_{ij}}{1-p_{ij}}\right)
= \beta_0 + \beta_1\text{Treatment}_{ij}
+ \beta_2(\text{Age}_{ij}-60) + u_j,
$$

where patient $i$ belongs to clinic $j$. The new term is

$$u_j \sim N(0, \sigma^2_{clinic}).$$

The parts play different roles:

- **Fixed intercept $\beta_0$:** average baseline log-odds for a 60-year-old
  standard-care patient in a typical clinic.
- **Fixed treatment effect $\beta_1$:** average adjusted treatment association
  across clinics.
- **Fixed age effect $\beta_2$:** average adjusted association for one more year
  of age.
- **Random intercept $u_j$:** clinic $j$'s deviation from the average baseline.
- **Random-effect SD $\sigma_{clinic}$:** how much baseline log-odds vary among
  clinics.

The model is called "mixed" because it contains both fixed and random effects.

## Fit the model

We use `glmer()` from **lme4**. Install the package once with
`install.packages("lme4")` if needed.

Conditional on a clinic effect $u_j$, the contribution of patient $i$ is

$$
P(Y_{ij}=y_{ij}\mid u_j)
=p_{ij}^{y_{ij}}(1-p_{ij})^{1-y_{ij}}.
$$

Because $u_j$ is unobserved, the mixed model integrates it out. The marginal
likelihood is

$$
L(\boldsymbol\beta,\sigma_{clinic})
=\prod_{j=1}^{J}
\int_{-\infty}^{\infty}
\left[\prod_{i=1}^{n_j}
p_{ij}(u)^{y_{ij}}\{1-p_{ij}(u)\}^{1-y_{ij}}
\right]
\phi(u;0,\sigma_{clinic}^2)\,du.
$$

`glmer()` numerically approximates these integrals while estimating the fixed
effects and clinic variance. The ordinary `glm()` is also fitted below so both
models use the identical outcome and predictors.

```{r fit-models}
if (!requireNamespace("lme4", quietly = TRUE)) {
  stop("This lesson requires the lme4 package. Run install.packages('lme4').")
}

ordinary_model <- glm(
  recovered ~ treatment + age_centered,
  data = medical,
  family = binomial
)

mixed_model <- lme4::glmer(
  recovered ~ treatment + age_centered + (1 | clinic),
  data = medical,
  family = binomial(link = "logit")
)

summary(mixed_model)
```

The notation `(1 | clinic)` means: allow every clinic to have its own
intercept, while estimating those clinic intercepts as draws from one shared
distribution.

## Interpret the fixed effects

As before, exponentiate fixed-effect coefficients to obtain odds ratios.

Within a clinic, the random intercept cancels from the treatment odds ratio:

$$
OR_{treatment\mid clinic}
=\frac{e^{\beta_0+\beta_1+\beta_2A+u_j}}
{e^{\beta_0+\beta_2A+u_j}}
=e^{\beta_1}.
$$

For fixed effect $k$, the cell calculates

$$
\widehat{OR}_k=e^{\widehat\beta_k},\qquad
CI_{95\%}=\left[
e^{\widehat\beta_k-1.96SE_k},
e^{\widehat\beta_k+1.96SE_k}
\right],
$$

and the two-sided Wald p-value

$$
p_k=2\Phi\left(-\left|\frac{\widehat\beta_k}{SE_k}\right|\right).
$$

```{r mixed-fixed-effects}
mixed_coefficients <- summary(mixed_model)$coefficients

mixed_results <- data.frame(
  term = rownames(mixed_coefficients),
  estimate_log_odds = mixed_coefficients[, "Estimate"],
  odds_ratio = exp(mixed_coefficients[, "Estimate"]),
  confidence_low = exp(mixed_coefficients[, "Estimate"] -
                         1.96 * mixed_coefficients[, "Std. Error"]),
  confidence_high = exp(mixed_coefficients[, "Estimate"] +
                          1.96 * mixed_coefficients[, "Std. Error"]),
  p_value = mixed_coefficients[, "Pr(>|z|)"],
  row.names = NULL
)

mixed_results
```

The treatment OR is a **clinic-specific**, or conditional, odds ratio: it
compares two otherwise similar patients in the same clinic. As in Lesson 1, the
p-value tests a no-association null hypothesis; it does not measure effect size,
clinical importance, or the probability that the null is true. Interpret it
with the OR, confidence interval, design, and clinical threshold.

The Wald intervals above are convenient teaching approximations. For important
research, consider likelihood-profile or bootstrap intervals and a pre-specified
analysis plan.

## Interpret the random effect

The fitted model estimates the clinic variance
$\widehat\sigma_{clinic}^2$. Its square root returns variation to the log-odds
scale:

$$
\widehat\sigma_{clinic}
=\sqrt{\widehat{\operatorname{Var}}(u_j)}.
$$

The following cell extracts these two quantities from `VarCorr()`.

```{r random-effect-variation}
clinic_variance <- as.data.frame(lme4::VarCorr(mixed_model))$vcov[1]
clinic_sd <- sqrt(clinic_variance)

c(clinic_variance = clinic_variance, clinic_sd = clinic_sd)
```

The clinic SD is measured on the log-odds scale. A value near zero suggests
little residual clinic-to-clinic variation after adjustment. A larger value
suggests that baseline recovery odds differ meaningfully among clinics.

For logistic mixed models, a commonly used latent-scale intraclass correlation
coefficient (ICC) is

$$
ICC \approx \frac{\sigma^2_{clinic}}
{\sigma^2_{clinic}+\pi^2/3}.
$$

```{r icc}
icc <- clinic_variance / (clinic_variance + pi^2 / 3)
icc
```

The ICC approximates the proportion of latent-scale variation attributable to
clinics. It is not literally the observed correlation between two binary
outcomes, so state the scale and approximation when reporting it.

## Clinic estimates and partial pooling

The estimated clinic deviations are:

For clinic $j$, `ranef()` returns the conditional mode

$$
\widehat u_j
=\arg\max_{u_j}
\left\{
\log P(\mathbf y_j\mid u_j,\widehat{\boldsymbol\beta})
+\log\phi(u_j;0,\widehat\sigma_{clinic}^2)
\right\}.
$$

The normal-density term penalizes extreme clinic effects. This is the
mathematical source of partial pooling: estimates with weak clinic-level
information are pulled more strongly toward $0$.

```{r clinic-random-intercepts}
clinic_estimates <- data.frame(
  clinic = rownames(lme4::ranef(mixed_model)$clinic),
  random_intercept = lme4::ranef(mixed_model)$clinic[, "(Intercept)"],
  row.names = NULL
)

clinic_estimates
```

Mixed models use **partial pooling**. A clinic's estimate is informed by its own
patients but pulled toward the overall mean, especially when the clinic has
little information. This is usually more stable than fitting an entirely
separate model for every clinic.

## Compare ordinary and mixed models

To compare like with like, define for each fitted model $m$

$$
\widehat{OR}_m=e^{\widehat\beta_{1m}},\qquad
SE_m=SE(\widehat\beta_{1m}),\qquad
p_m=2\Phi\left(-\left|\widehat\beta_{1m}/SE_m\right|\right).
$$

The helper function below applies these same equations to `glm` and `glmer`
objects and stacks the results.

```{r compare-models}
extract_treatment <- function(model) {
  model_table <- summary(model)$coefficients
  estimate <- model_table["treatment", "Estimate"]
  standard_error <- model_table["treatment", "Std. Error"]
  p <- model_table["treatment", "Pr(>|z|)"]

  c(
    odds_ratio = exp(estimate),
    confidence_low = exp(estimate - 1.96 * standard_error),
    confidence_high = exp(estimate + 1.96 * standard_error),
    standard_error = standard_error,
    p_value = p
  )
}

comparison <- rbind(
  ordinary_logistic = extract_treatment(ordinary_model),
  mixed_effects_logistic = extract_treatment(mixed_model)
)

comparison
```

The two estimates need not be identical. More importantly, the mixed model's
standard error acknowledges the grouping structure. Clustering often reduces
the effective amount of independent information, although the exact change
depends on the data and model. A p-value crossing 0.05 after accounting for
clustering does not mean the effect disappeared; it means the evidence and its
uncertainty were evaluated under a different, more suitable model.

Do not choose between these models based on which produces the smaller p-value.
Choose based on the study design and how the observations were generated.

## Population-average and clinic-specific predictions

A mixed model supports different prediction questions:

- **Typical-clinic prediction:** set the clinic random effect to zero. This is
  conditional on a clinic with average random intercept.
- **Known-clinic prediction:** include that clinic's estimated random effect.

Here are typical-clinic probabilities for a 60-year-old:

Setting the clinic deviation to its population mean, $u_j=0$, gives

$$
\widehat p_{typical}(t,a)
=\operatorname{logit}^{-1}
\left(\widehat\beta_0+\widehat\beta_1t+\widehat\beta_2a\right).
$$

In `predict()`, `re.form = NA` excludes all fitted random effects and implements
this equation. Setting `type = "response"` returns probability rather than
log-odds.

```{r typical-clinic-predictions}
new_patients <- data.frame(
  treatment = c(0, 1),
  age_centered = c(0, 0),
  clinic = factor(c(1, 1), levels = levels(medical$clinic))
)

new_patients$group <- c("Standard care", "New treatment")
new_patients$predicted_recovery_typical_clinic <- predict(
  mixed_model,
  newdata = new_patients,
  type = "response",
  re.form = NA
)

new_patients[, c("group", "predicted_recovery_typical_clinic")]
```

These are not fully marginal, population-averaged probabilities: simply setting
$u_j=0$ does not integrate over the distribution of clinics. That distinction
matters when selecting a prediction method for a clinical report.

## When should you use a mixed-effects model?

A mixed model is useful when observations are naturally grouped or repeated,
for example:

- patients within clinics or hospitals;
- repeated visits within patients;
- students within schools;
- measurements made by different clinicians or devices.

The grouping variable should represent a meaningful source of dependence. With
very few groups, random-effect variance and small-sample inference can be
unstable; alternatives may include clinic fixed effects, cluster-robust methods,
or small-sample corrections, depending on the research question and design.

Our random-intercept model assumes the treatment effect is the same across
clinics. If the treatment effect plausibly varies by clinic and enough clinics
and data are available, a random-slope model may be appropriate:

$$
\operatorname{logit}(p_{ij})
=\beta_0+\beta_1T_{ij}+\beta_2A_{ij}
+u_{0j}+u_{1j}T_{ij},
$$

$$
\begin{pmatrix}u_{0j}\\u_{1j}\end{pmatrix}
\sim N\left[
\begin{pmatrix}0\\0\end{pmatrix},
\begin{pmatrix}
\sigma_0^2 & \sigma_{01}\\
\sigma_{01} & \sigma_1^2
\end{pmatrix}
\right].
$$

Here $u_{0j}$ changes clinic $j$'s baseline, while $u_{1j}$ changes its treatment
effect. The R notation `(1 + treatment | clinic)` estimates both variation
terms and their correlation.

```{r random-slope-example, eval=FALSE}
random_slope_model <- lme4::glmer(
  recovered ~ treatment + age_centered + (1 + treatment | clinic),
  data = medical,
  family = binomial
)
```

That is a more demanding model and should be motivated by the design, not added
only because the syntax is available.

## Reporting template

> We fitted a mixed-effects logistic regression with fixed effects for treatment
> and age and a random intercept for clinic. The adjusted clinic-specific
> treatment odds ratio was **[OR]** (95% CI **[lower, upper]**, Wald p = **[p]**).
> The estimated between-clinic random-intercept SD was **[SD]**, corresponding
> to a latent-scale ICC of **[ICC]**. Effect magnitude, precision, clinical
> importance, and study limitations were considered together.

The reporting cell collects the mathematical summaries already derived:

$$
e^{\widehat\beta_1},\quad
e^{\widehat\beta_1\pm1.96SE_1},\quad
2\Phi(-|\widehat\beta_1/SE_1|),\quad
\widehat\sigma_{clinic},\quad
\frac{\widehat\sigma_{clinic}^2}
{\widehat\sigma_{clinic}^2+\pi^2/3}.
$$

```{r report-values}
treatment_result <- mixed_results[mixed_results$term == "treatment", ]

c(
  odds_ratio = treatment_result$odds_ratio,
  confidence_low = treatment_result$confidence_low,
  confidence_high = treatment_result$confidence_high,
  p_value = treatment_result$p_value,
  clinic_sd = clinic_sd,
  latent_scale_icc = icc
)
```

## Check your understanding

1. What dependence does `(1 | clinic)` represent?
2. Why is the treatment OR described as clinic-specific?
3. If the mixed-model p-value is larger than the ordinary-model p-value, did the
   treatment effect necessarily become smaller?
4. What does partial pooling do to clinic estimates?
5. When might a random treatment slope be scientifically justified?
